Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2022-02-07 move_schedules.tex
DC&G prelim reformat

\pdfoutput=1
% \documentclass[review]{siamart190516} % for SICOMP 
\documentclass{article} % for arxiv 
\usepackage{arxiv}

%\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{scalefnt}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{algorithm}
%\usepackage{algorithm2e}
%\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\usepackage{bbm}
\usepackage{graphicx}

\pagestyle{plain} % arxiv only

\newcommand\sbullet[1][.5]{\mathbin{\vcenter{\hbox{\scalebox{#1}{$\bullet$}}}}}
\usepackage{blkarray, bigstrut}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, calc, quotes, tikzmark,shapes}

@@

\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}}

\newcommand\topstrut[1][1.1ex]{\setlength\bigstrutjot{#1}{\bigstrut[t]}}
\newcommand\botstrut[1][0.95ex]{\setlength\bigstrutjot{#1}{\bigstrut[b]}}

@@

% For arxiv; comment out for SICOMP
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proof}{Proof}
% \newcommand{\CommentX}[1]{\unskip~\%~#1~\%}

@@

% Authors: full names plus addresses.
\author{
	Matthew Piekenbrock\thanks{Khoury College of Computer Sciences, Northeastern University.}
	\and Jose A. Perea\thanks{
	Department of Mathematics and Khoury College of Computer Sciences, Northeastern University.}
}

@@

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input, which motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	In practice, the quadratic scaling in the number of  such transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch,  limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we present results showing that the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
	Applications to multi-dimensional persistence and crocker stacks are also presented.
\end{abstract}

% REQUIRED
% See: https://mathscinet.ams.org/mathscinet/msc/msc2020.html?t=68T09&btn=Current
%\begin{AMS}
%  68T09, 55N31, 62R40
%\end{AMS}

@@

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we present applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

A VR filtration indexed over $\mathbb{R}$  can be condensed to an essential filtration indexed over the set of pairwise distances $\{\, d(x,x') \; | \; x,x' \in X \, \}$, 

@@

  		w \kern\rowsp & \; 1 &  \fbox{1} &    \; \botstrut \\

@@

  		v \kern\rowsp & \;     & 1 & \fbox{1} \; \\

@@

@@

	\,\,, 

@@

        a \kern\rowsp  & \; 1 &  \fbox{1} & 1 \; \topstrut \\

@@

  		 & \; 1 &  &  \; \topstrut \\
  		 & \;  & 1 &  \; \\
  		 & \; 1 & 1 & \; \botstrut \\
	\,\,, 

@@

  		& \;  & 1 &  \; \topstrut \\
  		& \; 1 &  &  \; \\
  		& \; 1 & \fbox{1} &  \; \botstrut \\
	\,\,, 

@@

  		u \kern\rowsp  & \;  & 1 &  \; \topstrut \\
  		v \kern\rowsp & \; 1 & 1 &  \; \\
	\,\,,
  		  & \; 1 & 1 & 1 \; \topstrut \\

@@

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition.
 Informally, since column operations are the largest complexity operations needed and each column can have potentially $m$ entries, the complexity of vineyards is $O(m)$ per transposition. 
 However, there are a number of subtleties involved in this statement. 
 The first  is that both the $V$ and $R$ matrices tend to be quite sparse, and thus column and row operations rarely require $m$ operations---in fact, as a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. A second subtlety, due to the necessity of sparse matrix data structures, is that querying the non-zero status of any particular entry typically takes $O(\log m)$ time. This hidden cost may seem minor, however, the quadratic scaling of the number of transpositions vineyards typically requires makes this cost non-trivial.   
In fact, achieving $O(m)$ complexity per transposition requires a special sparse matrix representation that allows swapping any two rows and columns in $O(1)$---see the appendix for more details and discussion. Finally, note the constant factor associated with the stated complexity: inspection of the individual cases of the algorihtm from~\cite{cohen2006vines} shows that there are at most two $O(m)$ operations, on both $R$ and $V$, that may be needed for any single transposition.

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[labelsep=5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in either $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid  during the movement of $\sigma_i $ to $ \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times.
By comparison, if a row-oriented sparse matrix implementation is used, then a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

@@

\noindent \textbf{Example:} We continue the example used in sections~\ref{sec:vineyards} and~\ref{sec:reduction} to illustrate moves. Consider moving the edge $a$ to the position of edge $c$ in the filtration. The donor columns introduced in~\cite{busaryev2010tracking} are shown on the left side of each matrix. Note that using vineyards, the equivalent permutation to the decomposition would require $4$ column operations on both $R_1$ and $V_1$, respectively, whereas a single move operation accomplishes this permutation using only $2$ column operations per matrix. 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.95}
	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
	%
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		u \kern \rowsp & \; 1 \; \topstrut \\
  		v \kern \rowsp & \;  \; \\
  		w \kern \rowsp & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{R_1}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern \rowsp & \; 1 & 1 &  \; \topstrut \\
  		v \kern \rowsp & \;  & 1 &  \; \\
  		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \;  1 \; \\
  		 & \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &    &  \; \topstrut \\
  		  & \;    & 1 &  \; \\
  		 & \; 1 & \fbox{1} &  \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \;  \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \; 
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		  & \; 1   &    &   1  \; \topstrut \\
  		 & \;       &  1  &  1  \; \\
  		 & \; 1   &   \fbox{1}   &     \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} 
\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \; \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		  & \;  & 1 &  1\; \topstrut \\
  		 & \; 1 & 1 &  \; \\
  		 & \; 1 &  & \fbox{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{d_R} 
\begin{blockarray}{cccc}{R_1}
\kern-0.2em & b & c & a  \\
	\begin{block}{c[ccc]}
		u \kern \rowsp & \;  & 1 &  \; \topstrut \\
		v \kern \rowsp & \; 1 & 1 &  \; \\
		w \kern \rowsp & \; 1 &  &  \; \botstrut \\
	\end{block}
	% \undermat{B}{ & & }  
\end{blockarray}
\end{displaymath}

\vspace{-1.0em}

% Start V example 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.95}
	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
	%
	\begin{blockarray}{cc}{d_V}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		 & \; 1 \; \topstrut \\
  		& \;  \; \\
  		& \;  \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{V_1}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		 & \; 1 & \fbox{1} &  \fbox{1}\; \topstrut \\
  		& \;  & 1 &  1\; \\
  		& \; &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		 & \; 1 \; \topstrut \\
  		& \;  1 \; \\
  		& \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		 & \; 1 & &  \fbox{1}\; \topstrut \\
  		& \;  & 1 &  1\; \\
  		& \; &   & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		 & \; 1 \; \topstrut \\
  		& \;  1 \; \\
  		& \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		 & \; 1 & &   \; \topstrut \\
  		& \;  & 1 &   \; \\
  		& \;  &   & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} \\
% 
\begin{blockarray}{cc}{}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		 & \; 1 \; \topstrut \\
  		& \;  1 \; \\
  		& \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		 & \; 1 & &\; \topstrut \\
  		& \;  & 1 &  \; \\
  		& \; &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{d_V}
% 
	\begin{blockarray}{cccc}{V_1}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		b \kern \rowsp & \; 1 & & 1 \; \topstrut \\
  		c \kern \rowsp & \; & 1 &  1 \; \\
  		a \kern \rowsp & \; &  & 1  \; \botstrut \\
		\end{block}
	\end{blockarray}
\end{displaymath}

@@

\scalefont{1.0}{

@@

\begin{tabular}{ |P{1.6cm}|P{1.6cm}|P{1.2cm}|P{2.8cm}|P{2.4cm}|}

@@

  \multicolumn{1}{|c|}{ $\beta(M)$ } & \multicolumn{1}{|c|}{$\mathcal{A}(M)$ } & \multicolumn{3}{c|}{ Column reductions / Permutations } \\

@@

% REQUIRED: siamplain format
%\bibliographystyle{siamplain}
\bibliographystyle{unsrt}
\bibliography{move_schedules}

>>>>

% \documentclass[review]{siamart190516} % for SICOMP 
%\documentclass{article} % for arxiv 
% \usepackage{arxiv}
%\pagestyle{plain} % arxiv only 
\documentclass[sn-mathphys]{sn-jnl}

%\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{scalefnt}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{algorithm}

\usepackage{array}
%\usepackage{algorithm2e}
%%\usepackage{algorithmicx}
%\usepackage[noend]{algpseudocode}
%\usepackage{bbm}
\usepackage{graphicx}

\usepackage{blkarray, bigstrut}
\newcommand\topstrut[1][1.2ex]{\setlength\bigstrutjot{#1}{\bigstrut[t]}}
\newcommand\botstrut[1][0.9ex]{\setlength\bigstrutjot{#1}{\bigstrut[b]}}

@@

@@

% For arxiv; comment out for SICOMP
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
% \newcommand{\CommentX}[1]{\unskip~\%~#1~\%}

@@

% Authors: full names plus addresses.
\author{
	Matthew Piekenbrock\thanks{Khoury College of Computer Sciences, Northeastern University.}
	\& Jose A. Perea\thanks{
	Department of Mathematics and Khoury College of Computer Sciences, Northeastern University.}
}

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract}
\abstract{
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input, which motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	In practice, the quadratic scaling in the number of  such transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch,  limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we present results showing that the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
	Applications to multi-dimensional persistence and crocker stacks are also presented.
}
%\end{abstract}

% REQUIRED
% See: https://mathscinet.ams.org/mathscinet/msc/msc2020.html?t=68T09&btn=Current
%\begin{AMS}
%  68T09, 55N31, 62R40
%\end{AMS}
\pacs[MSC Classification]{68T09, 55N31, 62R40}

@@

@@

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we present applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work.

@@

A VR filtration indexed over $\mathbb{R}$  can be condensed to an essential filtration indexed over the set of pairwise distances $\{\, d(x,x') \; \mid \; x,x' \in X \, \}$, 

@@

  		w \kern\rowsp & \; 1 &  \underline{1} &    \; \botstrut \\

@@

  		v \kern\rowsp & \;     & 1 & \underline{1} \; \\

@@

  		u \kern\rowsp  & \; 1 & 1 &  \hphantom{1} \; \topstrut \\
  		v \kern\rowsp & \;  & 1 &  \hphantom{1} \; \\
  		w \kern\rowsp & \; 1 &  &  \hphantom{1} \; \botstrut \\  

@@

  		 & \; 1 &  &  \hphantom{1} \; \topstrut \\
  		 & \;  & 1 &  \hphantom{1} \; \\
  		 & \; 1 & 1 & \hphantom{1} \; \botstrut \\
  		& \;  & 1 &  \hphantom{1} \; \topstrut \\
  		& \; 1 &  &  \hphantom{1} \; \\
  		& \; 1 & \underline{1} &  \hphantom{1} \; \botstrut \\
  		u \kern\rowsp & \;  & 1 &  \hphantom{1} \; \topstrut \\
  		v \kern\rowsp & \; 1 & 1 &  \hphantom{1} \; \\
  		w \kern\rowsp & \; 1 &  &  \hphantom{1} \; \botstrut \\
\vspace{-3.0em}
% Start V 

@@

        a \kern\rowsp  & \; 1 &  \underline{1} & 1 \; \topstrut \\

@@

	\, \,

@@

@@

@@

%	\def\rowsp{0.4em}
%	\def\arraystretch{0.95}
%	\makeatletter\setlength\BA@colsep{1.2pt}\makeatother
%	%
%	\begin{blockarray}{cccc}{R_1}
%	 \kern-0.2em & a & b & c  \\
%  		u \kern\rowsp  & \; 1 & 1 &  \; \topstrut \\
%  		v \kern\rowsp & \;  & 1 &  \; \\
%  		w \kern\rowsp & \; 1 &  &  \; \botstrut \\  
%	\,\,, 
%	\begin{blockarray}{cccc}{V_1}
%	\kern-0.2em & a & b & c  \\
%        a \kern\rowsp  & \; 1 &  \fbox{1} & 1 \; \topstrut \\
%  		b \kern\rowsp & \;  & 1 & 1 \; \\
%  		c \kern\rowsp & \;  &  & 1 \; \botstrut \\
%	\xrightarrow{S_{1}^2} 
%	\hspace{-0.1em}
%	\begin{blockarray}{cccc}{}
%	\kern-0.2em & a & b & c  \\
%  		 & \; 1 &  &  \; \topstrut \\
%  		 & \;  & 1 &  \; \\
%  		 & \; 1 & 1 & \; \botstrut \\
%	\,\,, 
%	\begin{blockarray}{cccc}{}
%	\kern-0.2em & a & b & c  \\
%  		 & \; 1 &  & 1 \; \topstrut \\
%  		 & \;  & 1 & 1 \; \\
%  		 & \;  &  & 1 \; \botstrut \\
%	\xrightarrow{P}
%	\hspace{-0.2em}
%	\begin{blockarray}{cccc}{}
%	\kern-0.2em & b & a & c  \\
%  		& \;  & 1 &  \; \topstrut \\
%  		& \; 1 &  &  \; \\
%  		& \; 1 & \fbox{1} &  \; \botstrut \\
%	\,\,, 
%	\begin{blockarray}{cccc}{}
%	\kern-0.2em & b & a & c  \\
%  		& \; 1 &  & 1 \; \topstrut \\
%  		& \;  & 1 & 1 \; \\
%  		& \;  &  & 1 \; \botstrut \\
%	\xrightarrow{S_1^2}
%	\hspace{-0.1em}
%	\begin{blockarray}{cccc}{}
%	\kern-0.2em & b & a & c  \\
%  		u \kern\rowsp  & \;  & 1 &  \; \topstrut \\
%  		v \kern\rowsp & \; 1 & 1 &  \; \\
%  		w \kern\rowsp & \; 1 &  &  \; \botstrut \\
%	\,\,,
%	\begin{blockarray}{cccc}{}
%	\kern-0.2em & b & a & c  \\
%  		  & \; 1 & 1 & 1 \; \topstrut \\
%  		 & \; & 1 & 1 \; \\
%  		 & \;  &  & 1 \; \botstrut \\

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition.
 Informally, since column operations are the largest complexity operations needed and each column can have potentially $m$ entries, the complexity of vineyards is $O(m)$ per transposition. 
 However, there are a number of subtleties involved in this statement. 
 The first  is that both the $V$ and $R$ matrices tend to be quite sparse, and thus column and row operations rarely require $m$ operations---in fact, as a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. A second subtlety, due to the necessity of sparse matrix data structures, is that querying the non-zero status of any particular entry typically takes $O(\log m)$ time. This hidden cost may seem minor, however, the quadratic scaling of the number of transpositions vineyards typically requires makes this cost non-trivial.   
In fact, achieving $O(m)$ complexity per transposition requires a special sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in~\cite{cohen2006vines}. Finally, note the constant factor associated with the stated complexity: inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that there are at most two $O(m)$ operations, on both $R$ and $V$, that may be needed for any single transposition.

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[labelsep=5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid  during the movement of $\sigma_i $ to $ \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times.
By comparison, if a row-oriented sparse matrix implementation is used, then a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

@@

\noindent \textbf{Example:} We continue the example used in sections~\ref{sec:vineyards} and~\ref{sec:reduction} to illustrate moves. Consider moving the edge $a$ to the position of edge $c$ in the filtration. The donor columns introduced in~\cite{busaryev2010tracking} are shown on the left side of each matrix. Note that using vineyards, the equivalent permutation to the decomposition would require $4$ column operations on both $R_1$ and $V_1$, respectively, whereas a single move operation accomplishes this permutation using only $2$ column operations per matrix. 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.60}
	\makeatletter\setlength\BA@colsep{0.9pt}\makeatother
	%
	\begin{blockarray}{cc}{d_R}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		u \kern \rowsp & \; 1 \; \topstrut \\
  		v \kern \rowsp & \;  \; \\
  		w \kern \rowsp & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{R}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		u \kern \rowsp & \; 1 & 1 &  \hphantom{1} \; \topstrut \\
  		v \kern \rowsp & \;  & 1 &  \hphantom{1} \; \\
  		w \kern \rowsp & \; 1 &  &  \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \;  1 \; \\
  		 & \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &    &  \hphantom{1} \; \topstrut \\
  		  & \;    & 1 &  \hphantom{1} \; \\
  		 & \; 1 & \underline{1} &  \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \;  \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \; 
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		  & \; 1   &    &   1  \; \topstrut \\
  		 & \;       &  1  &  1  \; \\
  		 & \; 1   &   \underline{1}   &     \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} 
\begin{blockarray}{cc}{}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		  & \; \hphantom{1} \; \topstrut \\
  		 & \; \hphantom{1} \; \\
  		 & \; \hphantom{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		  & \;  & 1 &  1\; \topstrut \\
  		 & \; 1 & 1 &  \; \\
  		 & \; 1 &  & \underline{1} \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\xrightarrow{d_R} 
\begin{blockarray}{cccc}
\kern-0.2em & b & c & a  \\
	\begin{block}{c[ccc]}
		& \;  & 1 &  \hphantom{1} \; \topstrut \\
		& \; 1 & 1 &  \hphantom{1} \; \\
		& \; 1 &  &  \hphantom{1} \; \botstrut \\
	\end{block}
	% \undermat{B}{ & & }  
\end{blockarray}
\end{displaymath}

\vspace{-3.0em}

% start V example (1) 
\begin{displaymath}
	\def\rowsp{0.4em}
	\def\arraystretch{0.60}
	\makeatletter\setlength\BA@colsep{0.9pt}\makeatother
	%
	\begin{blockarray}{cc}{d_V}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		a \kern \rowsp & \; 1 \; \topstrut \\
  		b \kern \rowsp & \;  \; \\
  		c \kern \rowsp & \;  \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
	\begin{blockarray}{cccc}{V}
	 \kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		a \kern \rowsp & \; 1 & \underline{1} & \underline{1} \; \topstrut \\
  		b \kern \rowsp & \;  & 1 & 1 \; \\
  		c \kern \rowsp & \;  &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
\begin{blockarray}{cc}{}
	\kern-0.2em & b \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \;  1 \; \\
  		 & \; \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b & c  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &    & \underline{1} \; \topstrut \\
  		  & \;    & 1 &  1 \; \\
  		 & \;  & &  1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	%
\; \to
	%
	\begin{blockarray}{cc}{}
	\kern-0.2em & c \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \; 1 \; \\
  		 & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
	\; \; 
	\begin{blockarray}{cccc}{}
	\kern-0.2em & a & b &  c \\
		\begin{block}{c[ccc]}
  		  & \; 1   &    &    \; \topstrut \\
  		 & \;   &  1  &   \; \\
  		 & \;   &   &  1   \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\; \xrightarrow{P} 
\begin{blockarray}{cc}{}
	\kern-0.2em & a \\
		\begin{block}{c[c]}
  		  & \; 1 \; \topstrut \\
  		 & \; 1 \; \\
  		 & \; 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
\; \;
\begin{blockarray}{cccc}{}
	\kern-0.2em & b & c & a  \\
		\begin{block}{c[ccc]}
  		  & \; 1 &  &  \; \topstrut \\
  		 & \;  & 1 &  \; \\
  		 & \; &  & 1 \; \botstrut \\
		\end{block}
	\end{blockarray}
%
\xrightarrow{d_V} 
\begin{blockarray}{cccc}
\kern-0.2em & b & c & a  \\
	\begin{block}{c[ccc]}
		& \;  1 & &  1 \; \topstrut \\
		& \;  & 1 &  1\; \\
		& \; &  & 1 \; \botstrut \\
	\end{block}
	% \undermat{B}{ & & }  
\end{blockarray}
\end{displaymath}

@@

\setlength\arraycolsep{1.8pt}
\scalefont{0.90}{

@@

\begin{tabular}{ m{1.6cm} m{1.6cm} m{1.2cm} m{2.8cm} m{2.4cm} }

@@

  \multicolumn{1}{|c|}{ $\beta(M)$ } & \multicolumn{1}{|c|}{$\mathcal{A}(M)$ } & \multicolumn{3}{c|}{ Col. Reductions / Permutations } \\

@@

% REQUIRED: siamplain format
%\bibliographystyle{siamplain}
%\bibliographystyle{unsrt}
\bibliography{move_schedules}
